An efficient composition of bidirectional programs by tupling and lazy updates

### 2020-02-21 paper/1.introduction.tex
fixed based on Prof. Hu's comment

\noindent
\textcolor{red}{Unlike function composition, the composition of bidirectional programs is read left-to-right. We use this order because it is standard in the BX community.}
%\textcolor{red}{Note that the order of composition is the reverse order compared to function composition. We use this order, because it is standard in BX community.}
One feature of this composition is that $put_{bx_1 \circ bx_2}$ needs to call $get_{bx_1}$ to compute the intermediate result for $put_{bx_2}$ to use, which would introduce an efficiency problem if we compute $put$ for composition of many bidirectional programs. Generally, for a composition of $O(n)$ bidirectional programs, we need to call $get$ for $O(n^2)$ times. To be concrete, consider the evaluation of the following composition (which will be used as our running example in this paper):

@@

To treat large inputs, we propose the second method based on memoization: ``keeping \emph{complements in a closure} and using them when needed''.
\textcolor{red}{Here, complements are minimal parts of sources that are needed to evaluate $put$. 
  In the middle $put$ and $get$ of Figure \ref{fig:eval-comp-phead},
  we use $i_1$, $[[1,2],[3]]$ and $i_2'$, $[100,2]$ to obtain the updated source $[[100,2],[3]]$. However, $[1,2]$ is simply replaced by $[100,2]$ and not used to construct the result. In this case we can use $[..., [3]]$ as a complement.}
The key idea is straightforward: Complements are smaller
% program fragments
than
% the original
intermediate states.
%Therefore this strategy solves the previous two problems.
%Readers might already know that it is a hard problem to obtain complements in general. In our work, thanks to the following finding, it is not hard:
%
%\vspace{2mm}
%In very-well behaved \cite{Foster:2007:CBT:1232420.1232424} BX,% programs,
%$put$ is a complement function for $get$.
%, $get$ can be a complement function for $put$.
%\vspace{2mm}
%
For obtaining complements, we tuple $put$ and $get$, and produce a new function $pg$. Because $put$ produces new complements for $get$, we can shrink the size.
Let us reconsider our example $phead \circ phead \circ phead$ in Figure \ref{fig:eval-comp-phead-2}, where $c_1$ and $c_2$ are complements and $d_1$ and $d_2$ are valid views for $s$ and~$i_1$. Here, two points are worth noting. First, after evaluating the leftmost $pg$, the original source $s$ need not be kept because its complete contents are in $c_1$ and~$i_1$. Second, the complements are smaller than the intermediate states in Figure~\ref{fig:eval-comp-phead}.
%This evaluation looks better than previous two strategies: this does not require repeated evaluation and require smaller storage than the original sources.
Actually, this simple
% combined
$pg$ alone is not yet effective for left associative compositions because it requires two more $put$s, which can be seen on the right side of Figure \ref{fig:eval-comp-phead-2}. To achieve an efficient evaluation, we combine two techniques, lazy update and lazy evaluation. We explain this second method and all optimizations in Section~\ref{sect:xpg}.

>>>>

\noindent
\textcolor{red}{Note that the order of composition is different from that of the common function composition. We use this order, because it is helpful to understand the behavior if we consider data goes from left to right.}
One feature of this composition is that $put_{bx_1 \circ bx_2}$ needs to call $get_{bx_1}$ to compute the intermediate result for $put_{bx_2}$ to use, which would introduce an efficiency problem if we compute $put$ for composition of many bidirectional programs. Generally, for a composition of $O(n)$ bidirectional programs, we need to call $get$ for $O(n^2)$ times. To be concrete, consider the evaluation of the following composition (which will be used as our running example in this paper):

@@

To treat large inputs, we propose the second method based on memoization: ``keeping \emph{complements in a closure} and using them when needed''.
\textcolor{red}{Here, complements are information from sources that makes $get$ injective, which is in turn needed to evaluate $put$.
  In the middle $put$ and $get$ of Figure \ref{fig:eval-comp-phead},
  we use $i_1$, $[[1,2],[3]]$ and $i_2'$, $[100,2]$ to obtain the updated source $[[100,2],[3]]$. However, $[1,2]$ is simply replaced by $[100,2]$ and not used to construct the result. In this case we can use $[..., [3]]$ as a complement.}
The key idea is straightforward: Complements are smaller
% program fragments
than
% the original
intermediate states.
%Therefore this strategy solves the previous two problems.
%Readers might already know that it is a hard problem to obtain complements in general. In our work, thanks to the following finding, it is not hard:
%
%\vspace{2mm}
%In very-well behaved \cite{Foster:2007:CBT:1232420.1232424} BX,% programs,
%$put$ is a complement function for $get$.
%, $get$ can be a complement function for $put$.
%\vspace{2mm}
%
For obtaining complements, we tuple $put$ and $get$, and produce a new function $pg$. Because $put$ produces new complements for $get$, we can shrink the size.
Let us reconsider our example $phead \circ phead \circ phead$ in Figure \ref{fig:eval-comp-phead-2}, where $c_1$ and $c_2$ are complements and $d_1$ and $d_2$ are valid views for $s$ and~$i_1$. Here, two points are worth noting. First, after evaluating the leftmost $pg$, the original source $s$ need not be kept because its complete contents are in $c_1$ and~$i_1$. Second, the complements are smaller than the intermediate states in Figure~\ref{fig:eval-comp-phead}.
%This evaluation looks better than previous two strategies: this does not require repeated evaluation and require smaller storage than the original sources.
Actually, this simple
% combined
$pg$ alone is not yet effective for left associative compositions because it requires two more $put$s, which can be seen on the right side of Figure \ref{fig:eval-comp-phead-2}. To achieve an efficient evaluation, we combine two techniques, lazy update and lazy evaluation. We explain this second method and all optimizations in Section~\ref{sect:xpg}.

### 2020-02-21 paper/4.xpg.tex
Tiny update
English polished.
fixed based on Prof. Hu's comment

    A restriction for $\pginline{Case}$ needs to be introduced here. We know that there is one entering condition and one exit condition when evaluating $\putbxinline{Case}$ as well as $\getbxinline{Case}$. If a tupling occurs, there will be 4 combinations from these conditions. This means two entering conditions of $\putbxinline{Case}$ and $\getbxinline{Case}$ are not always simultaneously satisfied. The evaluated branches are distinct in the $put$ and $get$ directions for combinations $((cond_{sv} \ s \ v) \ \&\& \ (not (cond_{s} \ s)))$ and $((not (cond_{sv} \ s \ v)) \ \&\& \ (cond_{s} \ s))$, which are restricted in this paper\footnote{\textcolor{red}{If the BX programs have efficient inverse way for evaluation, they are not affected by this restriction. Of course, all BX programs do not always have such efficient way. An example restricted by this is evaluating $bsnoc$ with different length of a source and a view.}}. This does not happen for the others which is used in the construction of $\pginline{Case}$.

@@

\textcolor{red}{
In $kpg$, basically, functions for the updates will be kept in $lks$ and $lkv$.
In $\kpginline{RearrS}$ and $\kpginline{RearrV}$, $f_1$ and $g_1$ are accumulated in $lks$ and $lkv$.
The kept functions are evaluated in $\kpginline{Skip}$ and $\kpginline{Case}$ by application of $lks \ s$ and $lkv \ v$. At the same time, the third argument and fourth argument are updated with the identity functions.
This evaluation is needed, because these definitions require the actual values, $es$ and $ev$.
Thanks to this accumulation in $kpg$, $\underline{lks_1}$ and $\underline{s_1}$ in $\kpginline{bx_1 \circ bx_2}$ are not evaluated as possible.}
%Except $\kpginline{Skip}$ and $\kpginline{Case}$ where $s$ and $v$ hold actual values, functions for the computation will be kept in $ks'$ and $kv'$.}

\textcolor{red}{Additionally we did two optimizations in $kpg$. The first is in $\kpginline{bx_1 \times bx_2}$.
  Because $es$ and $ev$ are not used in this definition, we do not need to evaluate.
  However, if we accumulate $lks$ and $lkv$, both might be evaluated independently in two assignments using $\kpginline{bx_1}$ and $\kpginline{bx_2}$. This includes the same computation. To remove duplicate evaluations, we evaluate actual values $es$ and $ev$ before calling $\kpginline{bx_1}$. and $\kpginline{bx_2}$}
%It is possible to evaluate actual values in $s$ and $v$ before calling $\kpginline{bx_1}$ to remove the redundancy like that. One more thing,
Second is in $\kpginline{Case}$ and not shown in the definition. We need to evaluate $lks' \ s'$ and $lkv' \ v'$ to check the \underline{fi} condition before returning the 6-tuple. Such evaluations can be done lazily to make programs run faster. We use the above small optimizations in our implementation.

>>>>

    A restriction for $\pginline{Case}$ needs to be introduced here. We know that there is one entering condition and one exit condition when evaluating $\putbxinline{Case}$ as well as $\getbxinline{Case}$. If a tupling occurs, there will be 4 combinations from these conditions. This means two entering conditions of $\putbxinline{Case}$ and $\getbxinline{Case}$ are not always simultaneously satisfied. The evaluated branches are distinct in the $put$ and $get$ directions for combinations $((cond_{sv} \ s \ v) \ \&\& \ (not (cond_{s} \ s)))$ and $((not (cond_{sv} \ s \ v)) \ \&\& \ (cond_{s} \ s))$, which are restricted in this paper. This does not happen for the others which is used in the construction of $\pginline{Case}$.
% {\textcolor{red}{If the BX programs have efficient inverse way for evaluation, they are not affected by this restriction. Of course, all BX programs do not always have such efficient way. An example restricted by this is evaluating $bsnoc$ with different length of a source and a view.}}

@@

\textcolor{red}{
In $kpg$, basically, functions for the updates are kept (but not evaluated) in $lks$ and $lkv$.
In $\kpginline{RearrS}$ and $\kpginline{RearrV}$, $f_1$ and $g_1$ are accumulated in $lks$ and $lkv$.
These two functions are evaluated in $\kpginline{Skip}$ and $\kpginline{Case}$ by applications of $lks \ s$ and $lkv \ v$. At the same time, the third arguments and fourth arguments of recursive calls are updated with the identity functions.
This evaluation is needed, because these definitions require the actual values, $es$ and $ev$.
Thanks to this update, accumulation in $kpg$, $\underline{lks_1}$ and $\underline{s_1}$ in $\kpginline{bx_1 \circ bx_2}$ are not evaluated as much as possible.}
%Except $\kpginline{Skip}$ and $\kpginline{Case}$ where $s$ and $v$ hold actual values, functions for the computation will be kept in $ks'$ and $kv'$.}

\textcolor{red}{Additionally we did two optimizations in $kpg$. The first is in $\kpginline{bx_1 \times bx_2}$.
  Because $es$ and $ev$ are not used in this definition, we do not need to evaluate.
  However, if we accumulate $lks$ and $lkv$, both might be evaluated independently in two assignments using $\kpginline{bx_1}$ and $\kpginline{bx_2}$. This includes the same computation. To remove duplicate evaluations, we evaluate actual values $es$ and $ev$ before calling $\kpginline{bx_1}$. and $\kpginline{bx_2}$}.
%It is possible to evaluate actual values in $s$ and $v$ before calling $\kpginline{bx_1}$ to remove the redundancy like that. One more thing,
The second is in $\kpginline{Case}$ and not shown in the definition. We need to evaluate $lks' \ s'$ and $lkv' \ v'$ to check the \underline{fi} condition before returning the 6-tuple. Such evaluations can be done lazily to make programs run faster. We use the above small optimizations in our implementation.

### 2020-02-21 paper/5.experiments.tex
modified based on the discussion with Bach-san
Very tiny update
Footnote 7: spacing reduced.
fixed based on Prof. Hu's comment

We have fully implemented and tested all methods% 
%
\footnote{All experiments on macOS 10.14.6, processor Intel Core i7 (2.6~GHz), RAM 16~GB 2400~MHz DDR4, OCaml 4.07.1. The OCaml runtime system options and garbage collection parameters are set as default.}$^,$%HACK!
%
\footnote{Anonymous \url{https://mega.nz/\#F!ec8y2ITa!tLOHKkHgs_7fjSI8m-O26Q}}
%
described in the previous sections. Our target language is untyped. Some dummies used for $pg$, $cpg$, $kpg$ and $xpg$ are replaced with the current updated views. This helps a program in the $put$ direction valid.

@@

\textcolor{red}{TODO-D4: emphasize overhead of memoization. it is not for free.}
Figure \ref{fig:result} 
% illustrates 
shows the evaluation times for each of the seven test cases using the three methods: $put$ in minBiGUL, $put_m$ in minBiGUL$_m$ and $xpg$.
We also did similar experiments with $pg$, $cpg$ and $kpg$, but their results are slower than the corresponding ones of $xpg$. The slowness was caused by the exponential number of $pg$ calls in the case of $pg$, redundant evaluations ($cpg$), and redundant overhead for constructing closures ($kpg$).
Therefore we simply omit these results in Figure \ref{fig:result}.
As we all know, $put$ works poorly for left associative compositions because of the number of reevaluated $get$s.
%The figure shows $put_m$ and $xpg$ are better than $put$ in some cases.
The left part of the figure contains tests using not-large inputs, and we see that $put_m$ is the fastest method for them. However if the input size is large enough, in the cases of the right part of the figure, $put_m$ will be slower quickly due to time for manipulating data in the table. At that time, $xpg$ is 
% proved to be
the most effective method.
\textcolor{red}{TODO-B2: add short evaluation result of memory.}

>>>>

We have fully implemented and tested all methods% 
%
\footnote{All experiments on macOS 10.14.6, processor Intel Core i7 (2.6~GHz), RAM 16~GB 2400~MHz DDR4, OCaml 4.07.1. The OCaml runtime system options and garbage collection parameters are set as default.}$^,$%HACK!
%
\footnote{\textcolor{red}{The implementation is available: } \url{https://github.com/k-tsushima/pgs}}
%
described in the previous sections. Our target language is untyped. Some dummies used for $pg$, $cpg$, $kpg$ and $xpg$ are replaced with the current updated views. This helps a program in the $put$ direction valid.

@@

Figure \ref{fig:result} 
% illustrates 
shows the evaluation times for each of the seven test cases using the three methods: $put$ in minBiGUL, $put_m$ in minBiGUL$_m$ and $xpg$.
We also did similar experiments with $pg$, $cpg$ and $kpg$, but their results are slower than the corresponding ones of $xpg$. The slowness was caused by the exponential number of $pg$ calls in the case of $pg$, redundant evaluations ($cpg$), and redundant overhead for constructing closures ($kpg$).
Therefore we simply omit these results in Figure \ref{fig:result}.
As we all know, $put$ works poorly for left associative compositions because of the number of reevaluated $get$s.
%The figure shows $put_m$ and $xpg$ are better than $put$ in some cases.
The left part of the figure contains tests using not-large inputs, and we see that $put_m$ is the fastest method for them. However if the input size is large enough, in the cases of the right part of the figure, $put_m$ will be slower quickly due to time for manipulating data in the table. At that time, $xpg$ is 
% proved to be
the most effective method.

\textcolor{red}{Note that this result focuses on the BX programs that include many compositions. If the number of compositions is small in the BX programs, the original $put$ without any memoization will be fastest, because of the overhead for memoization ($put_m$) and the overhead for keeping complements in closures ($xpg$).}

%\textcolor{red}{TODO-B2: add short evaluation result of memory.}

### 2020-02-21 paper/6.related-work.tex
fixed based on Prof. Hu's comment

Since the pioneering work of lens \cite{Foster:2007:CBT:1232420.1232424}, many BX languages have been proposed \cite{Bohannon06relationallenses:, Bohannon:2008:BRL:1328438.1328487,Buchmann:2018:BFI:3362232.3362263,Ko:2016:BFV:2847538.2847544,LeblebiciAS14,Samimi-Dehkordi18,Cicchetti2011,Hinkel:2019:CPB:3318595.3318634}.
%(memo: BXtend, eMoflon, EVL+Strace, JTL, NMF. SDMLib should be added?)
%based on several application scenarios and BX lenses works \cite{}.
Although much progress has been made on the semantics and correctness of BX programs for the past years, as far as we are aware, little work has been done on optimization of BX programs \cite{Horn:2018:IRL:3243631.3236769}. Anjorin et. al. introduces the first benchmark for BX languages and compared them \cite{Anjorin2019}, but a systematic improvement for practical implementation of BX languages is still missing. This paper shows the first attempt of improving efficiency of BX composition evaluation.

>>>>

Since the pioneering work of lens \cite{Foster:2007:CBT:1232420.1232424}, many BX languages have been proposed \cite{Bohannon06relationallenses:, Bohannon:2008:BRL:1328438.1328487,Buchmann:2018:BFI:3362232.3362263,Ko:2016:BFV:2847538.2847544,LeblebiciAS14,Samimi-Dehkordi18,Cicchetti2011,Hinkel:2019:CPB:3318595.3318634}.
%(memo: BXtend, eMoflon, EVL+Strace, JTL, NMF. SDMLib should be added?)
%based on several application scenarios and BX lenses works \cite{}.
Although much progress has been made on the semantics and correctness of BX programs for the past years, as far as we are aware, little work has been done on optimization of BX programs \cite{Horn:2018:IRL:3243631.3236769}. Anjorin et. al. introduces the first benchmark\footnote{\textcolor{red}{The BX programs in their benchmark are basically BX programs without composition. Because we focus on the BX programs that include many compositions, their benchmark is not applicable for our evaluation.}} for BX languages and compared them \cite{Anjorin2019}, but a systematic improvement for practical implementation of BX languages is still missing. This paper shows the first attempt of improving efficiency of BX composition evaluation.

### 2020-02-21 paper/main.tex
change anonymus -> our name
tiny update main
(RG email updated)
fixed based on Prof. Hu's comment

@@

\author{{Kanae Tsushima\inst{1}\orcidID{0000-0002-3383-3389}} \and \break
  {Bach Nguyen Trong\inst{1}\orcidID{0000-0002-6342-6020}} \and \break
  {Robert Gl\"{u}ck\inst{2}\orcidID{0000-0001-6990-3935}} \and \break
  {Zhenjiang Hu\inst{3}\orcidID{0000-0002-9034-205X}}}

@@

\authorrunning{anonymous authors}

@@

\institute{National Institute of Informatics, Japan
  \email{\{k\_tsushima,bach\}@nii.ac.jp} \and
  University of Copenhagen, Denmark
  \email{glueck@di.ku.dk } \and
  Peking University, China
  \email{huzj@pku.edu.cn}}

>>>>

\newcommand{\newcirc}{\tilde{\circ}}

@@

\author{{Kanae Tsushima\inst{1}} \and \break% \orcidID{0000-0002-3383-3389}
  {Bach Nguyen Trong\inst{1}} \and \break %\orcidID{0000-0002-6342-6020}
  {Robert Gl\"{u}ck\inst{2}} \and \break %\orcidID{0000-0001-6990-3935}
  {Zhenjiang Hu\inst{3}}} %\orcidID{0000-0002-9034-205X}

@@

%\authorrunning{anonymous authors}

@@

\institute{National Institute of Informatics, Japan 
  (\email{\{k\_tsushima,bach\}@nii.ac.jp}) \and
  University of Copenhagen, Denmark 
  (\email{glueck@di.ku.dk}) \and
  Peking University, China
  (\email{huzj@pku.edu.cn})}
